Storm surges in the Strait of Georgia simulated with a regional model

### 2015-08-06 AO_guides/manuscriptAO_final.tex
Removed red text from final version draft

Coastal communities can prepare for and respond to storm surge hazards using predictions from storm surge models to determine if flooding conditions are probable. Like coastal ocean models, some of the complexity of a storm surge model depends on the properties of the geographic location of interest. In shallow regions, it is important to include the effects of wetting and drying which allows grid cells to alternate between land and water depending on the modelled sea surface height and depth \citep{hubbert1999storm, weisberg2006hurricane}.  Other regions require modelling over a very large domain since non-local effects from the open sea may influence surge propagation \citep{mercer2002barotropic, weisberg2006hurricane, lane2009verification}. Such large domains typically use nested or unstructured grids to resolve the details of complicated coastlines. Appropriate treatment of rivers \citep{flather1994storm}, accurate bathymetry and the inclusion {\color{red} of} surface waves \citep{xu2010storm} have also been identified as important.  Most of these models neglect the effects of stratification by treating the ocean as a constant temperature and salinity. An accurate forecast of the meteorological conditions at the ocean surface is of course essential in the generation of useful storm surge forecasts.

@@

%overview oF SoG/Salish Sea models
The early modelling work of \citet{crean1988mathematical}, \citet{crean1988numerical} and \citet{stronach1993update} {\color{red} developed} a series of models covering the Strait of Juan de Fuca, the Strait of Georgia, and Puget Sound using a finite difference discretization of the equations of motion. %A barotropic version of their model was used to reproduce storm surge water level elevations with reasonable accuracy, however, this barotropic version did not include the effects of fresh water entering the system from rivers. Later, a baroclinic version of their model examined residual circulation and fresh water flushing rates \citep{stronach1993update, marinone1996model}.
These models pointed to an important connection between the model's bottom friction parametrization and the location and amplitude of the M$_2$ amphidrome. They found that a local increase in the bottom friction coefficient, typically from 0.002 to 0.004 in the baroclinic model, especially in the vicinity of Haro Strait, was necessary in order to adequately reproduce the observed M$_2$ tidal amplitudes and phase lags in the Strait of Georgia \citep{stronach1993update}. Without local adjustments to the bottom friction their M$_2$ amplitudes were too low and the M$_2$ tidal phases were too large in the Strait of Georgia \citep{crean1988mathematical}. Local adjustments to bottom friction were applied in both their barotropic and baroclinic models.  Other studies over a similar domain, such as the barotropic finite element model of \citet{foreman1995tidal}, have avoided local increases in bottom friction although their domain did not include Puget Sound and the northern Strait of Georgia. \citet{foreman2004m} assimilated tide gauge harmonics to study M$_2$ tidal dissipation without local changes to the bottom friction in a two-dimensional finite element model that did not include baroclinic effects.

More recently, modelling studies have focused on baroclinic features. \citet{masson2004modelling} used the Princeton Ocean Model (POM) in a regional mode to examine the seasonal variability of deep water properties in the Strait of Georgia and noted a strong sensitivity to vertical mixing over the sills. This model employed a sigma or terrain-following vertical coordinate system leading to spurious diapycnal mixing since, numerically, the lateral mixing is computed on sigma levels which may intersect isopycnals in the presence of steep topography. 
%A finite volume model on an unstructured grid by \citet{yang2010multi} involved a truncated domain with an open boundary in the northern Strait of Georgia, where under predictions in the M$_2$ tidal phase over most of their domain were suggested to be caused by an inaccurate representation of the tides at their northern boundary. 
\citet{sutherland2011model} employed the Regional Ocean Modelling System (ROMS) to study estuarine circulation in the Salish Sea with a model domain that extended outside the Strait of Juan de Fuca into the Pacific Ocean and up to the northern Strait of Georgia. They noted that their semi-diurnal amplitudes were generally too low. {\color{red} An updated version of this model with added biogeochemical components and an extended domain produced higher than observed surface temperatures in the Strait of Juan de Fuca, suggesting too little mixing within the Salish Sea \citep{giddings2014hindcasts}.} Smaller domains have also been considered; notably, the passageways through the Discovery Islands north of the Strait of Georgia have been modelled baroclinically by \citet{foreman2012circulation} with the Finite Volume Coastal Ocean Model (FVCOM) pointing to difficulties in reproducing baroclinic currents on their grid.  Accurate reproduction of observed tidal amplitudes and phases and appropriate treatment of mixing is a recurring difficulty in modelling this region.

This article presents the development of a three-dimensional numerical ocean model for the Strait of Georgia and its skill in simulating storm surge hindcasts. {\color{red} The main goals are to identify the forcing factors that contribute most significantly to storm surges in the Strait of Georgia, to examine the spatial distribution of the storm surges, and to evaluate the performance of a new model of this region.}  To complement previous barotropic storm surge modelling in this region, this model employs a higher resolution and includes the effect of stratification and rivers. However, wetting and drying is not represented. Model evaluation has focused on regions where the near-coast slopes are steep and so wetting and drying effects are believed to be unimportant. A detailed overview of the model configuration and domain is provided in section \ref{sec:config}. That section also describes open and surface boundary conditions as well as choices in grid structure and mixing parametrizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's skill in producing storm surge hindcasts is assessed in section \ref{sec:storm}. Several simulations comparing the effects of different forcing conditions are also discussed. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

@@

We have used the Nucleus for European Modelling of the Ocean (NEMO) framework in its regional configuration \citep{madec2012nemo} to develop an ocean model for the Salish Sea. NEMO is a highly modular tool used for studying ocean physics, ocean-ice interactions, and the bio-geochemical properties of the ocean. NEMO's ocean core solves the three-dimensional hydrostatic equations of motion for an incompressible fluid under the Boussinesq approximation on a structured computational grid. {\color{red} Although these features are not being used in the present work, NEMO's options for grid nesting and bio-geochemical coupling make it a useful tool for studying the complex physics and bio-geochemical interactions within the Strait of Georgia.} This work focuses on the physical configuration of the Salish Sea model, in particular, determining appropriate forcing and boundary conditions for accurate reproduction of tidal amplitudes and phases as well as storm surge elevations. Future work will include bio-geochemical coupling and data assimilation. NEMO version 3.4 was used in this study. {\color{red} NEMO has been successfully used in other coastal modelling studies in the Northeast Atlantic \citep{maraldi2013nemo} and the Mediterranean Sea \citep{brossier2012sensitivity}.}

The modelled domain extends from the Strait of Juan de Fuca to Puget Sound to Johnstone Strait (Figure \ref{fig:domain}). Bathymetry from the Cascadia physiography data set \citep{haugerud1999digital} was smoothed to limit large changes in depth across grid cells, so that $\Delta h/\bar{h} \leq 0.8$, where $\Delta h$ is the difference in depth between two adjacent grid cells and $\bar{h}$ is their average depth. For model stability, additional smoothing at the Strait of Juan de Fuca western boundary was imposed to achieve constant depth across the first ten grid cells {\color{red} inward from the open boundary. Depths between 0 m and 4 m were set to 4 m.}  As depicted in Figure \ref{fig:domain}, the numerical grid is rotated $29^{\circ}$ counter-clockwise of North in order to maintain computational efficiency since currents within the Strait of Georgia are mainly aligned with this rotated axis. {\color{red} A magnified image of the complex passages through the San Juan and Gulf Islands is provided in Figure \ref{fig:zoom}.}

The curvilinear orthogonal numerical grid has been divided into 398 by 898 by 40 grid cells, which results in an almost uniform horizontal resolution with grid spacing approximately 440 m by 500 m. The 40 vertical $z$-levels were stretched gradually in order to achieve higher resolution in the surface layer, with 1 m vertical grid spacing down to about 10 m in depth. Below 10 m the grid was stretched to a maximum grid spacing of 27 m in the lowest layer. At the bottom boundary, partial $z$-levels were used in order to limit large changes in bathymetry across grid cells \citep{madec2012nemo}. {\color{red} Model constraints require that all passages be at least two grid points wide.  This restriction required choices in either widening or closing narrow straits.  In particular, Active Pass was closed and Porlier Pass was widened (Figure \ref{fig:zoom}). All of these compromises are expected to have significant effects locally but minimal effects on the large scale.}

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{Figures/bathy.pdf}
\caption{{\color{red} Model domain (highlighted area) including bathymetry, rivers (green circles), and storm surge locations of interest  (red stars). The black rectangle represents the region displayed in Figure \ref{fig:zoom}.}}\label{fig:domain}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[scale=0.4]{Figures/bathy_zoom.pdf}
\caption{{\color{red} Model bathymetry and coastlines in the passages between the San Juan and Gulf Islands.}}\label{fig:zoom}
\end{figure}

@@

Temperature and salinity at the Juan de Fuca boundary were taken from a weekly climatology which was created from results of a model covering the Salish Sea and the west coast of Vancouver Island \citep{massonfine2012}.  Their results, originally on $s$-levels were interpolated onto $z$-levels and then onto the NEMO horizontal grid to give a two-dimensional forcing that varies with time.  To prepare the climatology all years (1995-2008) were averaged and results, approximately every 15 days, were interpolated to a weekly climatology. The model is relaxed to the forced temperature and salinity over the 10 grid points (about 5 km) closest to the open boundaries, using a flow relaxation scheme \citep{engedahl1995use}. {\color{red} In the north, the seasonal climatologies of \citet{thomson1981oceanography} were used in the same manner.} %other references?

Sea surface height anomaly (or remote forcing) was set uniformly across the mouth of Juan de Fuca using values from the Tofino tide gauge \citep{DFOObservations} which is outside of, but close to, the western boundary of the model domain (Figure \ref{fig:domain}).  A monthly climatology was produced using daily averages from 2000-2010, binning them by month, averaging and setting the yearly mean to zero.  For the storm surge simulations, hourly variations in sea surface height were used.  These values are the Tofino tide gauge values with the tides removed. {\color{red}Tidal predictions were generated with the tidal analysis package called t\_tide \citep{pawlowicz2002classical} using a time series from the year of the storm surge. A harmonic analysis was performed for 67 constituents. In order to prevent over-predicted constituents due to high surge years, the time series was first filtered by applying a tide-killing Doodson filter \citep{parker2007NOAA} and periods with large non-tidal amplitudes were removed. Long period constituents such as Sa, Ssa, etc were not used in the tidal predictions since these constituents are often contaminated by non-tidal energy from seasonal meteorological trends. Further, constituents with a low signal to noise ratio were neglected. } The storm surge simulations also used the sea surface height anomaly from Port Hardy, forced at the northern boundary in Johnstone Strait.

{\color{red} Although Tofino is not the closest tide gauge to the western boundary, we found that the surge conditions on the western coast of Vancouver Island and near the mouth of the Strait of Juan de Fuca are fairly uniform as summarized in Table \ref{tab:pacificsurge}. Generally, surge amplitudes at Tofino are slightly larger and earlier than Neah Bay and Bamfield. The remote forcing is applied uniformly across the open boundaries. Given the similarities between the surge amplitude and phase at Neah Bay and Bamfield, this seems to be a reasonable assumption. In support of this assumption, a geostrophic slope calculation with an estuarine velocity of 20 cm s$^{-1}$ would imply a difference of less than 5 cm between the northern and southern sides of the Strait of Juan de Fuca. In the future, this uniform specification of sea level at the western boundary will be relaxed by taking values from the solution of a large-scale forecasting model.}

\begin{table}[h]
\centering 
\tbl{{\color{red}Average peak surge amplitudes and delay in peak timing from Tofino at several tide gauges on the Pacific coast. Averages obtained over the five surge events hindcasted in this article. A negative delay indicates that the peak surge occurred later than at Tofino.}} 
{\begin{tabular}{|l |c | c|}
\hline 
Location & Mean surge amplitude (m) & Mean delay from Tofino (hrs) \\
\hline
Port Hardy & 0.491 & -2.67\\
Winter Harbour & 0.572 & 0.333 \\
Tofino  & 0.707 & -- \\
Bamfield & 0.658 & -0.5 \\
Neah Bay & 0.658 & -1.83 \\
Toke Point & 1.17 & -2.83 \\
\hline
\end{tabular}}
\label{tab:pacificsurge}
\end{table}

The tidal forcing and {\color{red} non-tidal} remote forcing were applied using the NEMO Flather scheme \citep{flather1994storm, madec2012nemo}, a radiation scheme that allows disturbances to propagate out of the domain without reflections off of the open boundary. The barotropic velocity component of the remote forcing was set to zero. At the west, the baroclinic velocities at the boundary were set equal to the values inside the boundary (zero-gradient boundary conditions).  This scheme is not part of core NEMO 3.4 (the version used here).  Zero gradient conditions were chosen because the baroclinic velocity at the mouth of Juan de Fuca is primarily estuarine and thus set by density variations between inside and outside the domain. {\color{red} At the northern boundary, we choose the `no boundary condition' of NEMO that sets up an estuarine circulation by relaxing the temperature and salinity over a width of 10 grids to the seasonal climatology. }

@@

{\color{red} The ocean surface is forced with momentum fluxes, solar and non-solar heat fluxes, and precipitation from a 33-km global atmospheric reforecasting model {\color{red}called CGRF} suitable for use in ocean modelling \citep{smith2014new}.} Forecasts from the period of 2002-2012 are available. One additional simulation employs the western Canada component of the High Resolution Deterministic Prediction System (HRDPS), a nested 2.5 km resolution atmospheric model provided by Environment Canada \citep{ECModel}. In addition to momentum and heat fluxes, forcing due to the surface atmospheric pressure is also included in the ocean momentum equations \citep{madec2012nemo}. In the open ocean, the surface atmospheric pressure primarily induces an inverse barometer (IB) sea surface height,
\begin{equation}
 \eta_{IB} = -\frac{1}{g\rho_{0}}\left(P_{s}-P_0\right), \label{eq:inverse}
\end{equation}
where $g$ is the acceleration due to gravity, $\rho_{0}$ is a reference density set to 1035 kg m$^{-3}$, $P_{s}$ is the sea level atmospheric pressure from the atmospheric model, and $P_0$ is a reference sea level pressure (set to 101,000 Pa). Since the atmospheric model employs a terrain following vertical coordinate the model surface pressure has been corrected to sea level using \citep{holton1992introduction}
\[ P_s = P_1\left(\gamma\frac{z_1}{T_1} +1 \right)^\frac{g}{\gamma R},\]
where $P_1$ and $T_1$ are the model pressure and temperature at height $z_1$, $R$ is the ideal gas constant, and $\gamma$ is a temperature lapse rate set to 0.0098 K m$^{-1}$. In coastal and shelf seas, the atmospheric pressure forcing can also induce significant non-IB sea surface height, as demonstrated in previous studies of the Hudson Bay and Labrador shelf system \citep{wright1987influence,young1995synoptic}.

River input provides a significant volume of fresh water to the Salish Sea and can influence stratification, circulation, and primary productivity. However, most rivers in the domain are not gauged so parametrizations were required to represent river flow. \citet{morrison2011rivers} provides a method for estimating freshwater runoff in the Salish Sea region based on precipitation. Monthly runoff volumes for each watershed for each year from 1970 to 2012 were acquired from \citet{morrison2011rivers}, as well as monthly averages. Freshwater runoff from each watershed was divided among the rivers in that watershed. The area drained by each river was estimated from Toporama maps by the Atlas of Canada and watershed maps available on the Washington State government website. The watersheds included in our model were Fraser (which represents approximately 44\% of the freshwater input into our domain), Skagit (12\%), East Vancouver Island (North and South) (12\%), Howe (7\%), Bute (7\%), Puget (6\%), Juan de Fuca (5\%), Jervis (4\%) and Toba (3\%). The monthly flow from each river was input as a point source from 0 to 3 metres at the model point closest to the mouth of each river. Incoming water was assumed to be fresh and at surface {\color{red}ocean temperature}. A total of 150 rivers were parametrized by this method and their locations are indicated by the green dots in Figure \ref{fig:domain}.

Initial conditions for temperature and salinity came from a CTD cast in the middle Strait of Georgia taken in Sept 2002 \citep{pawlowiczetal2007}. Conditions were initially uniform horizontally and velocity was initialized at zero. The model was spun up for 15.5 months from the initial conditions above, starting Sept 16, 2002, using atmospheric forcing from 2002-2003, climatological temperature and salinity and sea surface height at the boundaries, with tides and climatological river output.  All storm surge runs were started at least three days prior to the event of interest with zero initial velocities and sea surface height and a stratification profile from model spin up. {\color{red} The modelled sea surface height adjusted to forcing in less than one day as determined by comparing a simulation with restart conditions to one with velocities and sea surface height initialized to zero}.

@@

In general, the K$_1$ tidal harmonics are well-represented by the model with complex differences below 5 cm at all stations except Seymour Narrows (station 27). Seymour Narrows is a very constricted region in Johnstone Strait and is not well resolved by the model. As such, the M$_2$ amplitude and phase are also poorly reproduced here.  The model's M$_2$ amphidrome, indicated by the drop in amplitude around station 5 (Victoria), is too strong resulting in lower than observed M$_2$ amplitudes in this region. {\color{red} The model also represents the dip in M$_2$ amplitude around stations 25 and 26 reasonably well.} Additionally, the M$_2$ phase errors in the Strait of Juan de Fuca (stations 1-7) are quite large, leading to large complex differences in this region.  The match in the Strait of Georgia (stations 10-24) is very good, with complex differences less than 6 cm, in part due to tidal tuning for highly accurate tides in the main region of interest.  {\color{red}Because the tides were tuned for a good match in the Strait of Georgia, the match in the Strait of Juan de Fuca was compromised; the effect is particularly noticeable at the station closest to the open boundary (Port Renfrew, 1). } Details on the tidal tuning procedure are provided in Appendix \ref{sec:appendix}.

@@

\section{Storm Surge Hindcasts}\label{sec:storm}
A series of five storm surge hindcasts is presented next. Storm surge simulations were started from rest at least three days prior to the event of interest to allow for adjustment of model currents. Stratification conditions from model spin up at a time close to the event (within five days) were chosen to initialize the temperature and salinity fields. {\color{red} Boundary conditions include atmospheric forcing (momentum and heat fluxes and precipitation) and fresh water input from rivers at the surface, sea surface height anomaly from observations at the open boundary, and tidal forcing with eight tidal constituents.}

>>>>

Coastal communities can prepare for and respond to storm surge hazards using predictions from storm surge models to determine if flooding conditions are probable. Like coastal ocean models, some of the complexity of a storm surge model depends on the properties of the geographic location of interest. In shallow regions, it is important to include the effects of wetting and drying which allows grid cells to alternate between land and water depending on the modelled sea surface height and depth \citep{hubbert1999storm, weisberg2006hurricane}.  Other regions require modelling over a very large domain since non-local effects from the open sea may influence surge propagation \citep{mercer2002barotropic, weisberg2006hurricane, lane2009verification}. Such large domains typically use nested or unstructured grids to resolve the details of complicated coastlines. Appropriate treatment of rivers \citep{flather1994storm}, accurate bathymetry and the inclusion  of surface waves \citep{xu2010storm} have also been identified as important.  Most of these models neglect the effects of stratification by treating the ocean as a constant temperature and salinity. An accurate forecast of the meteorological conditions at the ocean surface is of course essential in the generation of useful storm surge forecasts.

@@

%overview oF SoG/Salish Sea models
The early modelling work of \citet{crean1988mathematical}, \citet{crean1988numerical} and \citet{stronach1993update}  developed a series of models covering the Strait of Juan de Fuca, the Strait of Georgia, and Puget Sound using a finite difference discretization of the equations of motion. These models pointed to an important connection between the model's bottom friction parametrization and the location and amplitude of the M$_2$ amphidrome. They found that a local increase in the bottom friction coefficient, typically from 0.002 to 0.004 in the baroclinic model, especially in the vicinity of Haro Strait, was necessary in order to adequately reproduce the observed M$_2$ tidal amplitudes and phase lags in the Strait of Georgia \citep{stronach1993update}. Without local adjustments to the bottom friction their M$_2$ amplitudes were too low and the M$_2$ tidal phases were too large in the Strait of Georgia \citep{crean1988mathematical}. Local adjustments to bottom friction were applied in both their barotropic and baroclinic models.  Other studies over a similar domain, such as the barotropic finite element model of \citet{foreman1995tidal}, have avoided local increases in bottom friction although their domain did not include Puget Sound and the northern Strait of Georgia. \citet{foreman2004m} assimilated tide gauge harmonics to study M$_2$ tidal dissipation without local changes to the bottom friction in a two-dimensional finite element model that did not include baroclinic effects.

More recently, modelling studies have focused on baroclinic features. \citet{masson2004modelling} used the Princeton Ocean Model (POM) in a regional mode to examine the seasonal variability of deep water properties in the Strait of Georgia and noted a strong sensitivity to vertical mixing over the sills. This model employed a sigma or terrain-following vertical coordinate system leading to spurious diapycnal mixing since, numerically, the lateral mixing is computed on sigma levels which may intersect isopycnals in the presence of steep topography. 
%A finite volume model on an unstructured grid by \citet{yang2010multi} involved a truncated domain with an open boundary in the northern Strait of Georgia, where under predictions in the M$_2$ tidal phase over most of their domain were suggested to be caused by an inaccurate representation of the tides at their northern boundary. 
\citet{sutherland2011model} employed the Regional Ocean Modelling System (ROMS) to study estuarine circulation in the Salish Sea with a model domain that extended outside the Strait of Juan de Fuca into the Pacific Ocean and up to the northern Strait of Georgia. They noted that their semi-diurnal amplitudes were generally too low.  An updated version of this model with added biogeochemical components and an extended domain produced higher than observed surface temperatures in the Strait of Juan de Fuca, suggesting too little mixing within the Salish Sea \citep{giddings2014hindcasts}. Smaller domains have also been considered; notably, the passageways through the Discovery Islands north of the Strait of Georgia have been modelled baroclinically by \citet{foreman2012circulation} with the Finite Volume Coastal Ocean Model (FVCOM) pointing to difficulties in reproducing baroclinic currents on their grid.  Accurate reproduction of observed tidal amplitudes and phases and appropriate treatment of mixing is a recurring difficulty in modelling this region.

This article presents the development of a three-dimensional numerical ocean model for the Strait of Georgia and its skill in simulating storm surge hindcasts.  The main goals are to identify the forcing factors that contribute most significantly to storm surges in the Strait of Georgia, to examine the spatial distribution of the storm surges, and to evaluate the performance of a new model of this region.  To complement previous barotropic storm surge modelling in this region, this model employs a higher resolution and includes the effect of stratification and rivers. However, wetting and drying is not represented. Model evaluation has focused on regions where the near-coast slopes are steep and so wetting and drying effects are believed to be unimportant. A detailed overview of the model configuration and domain is provided in section \ref{sec:config}. That section also describes open and surface boundary conditions as well as choices in grid structure and mixing parametrizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's skill in producing storm surge hindcasts is assessed in section \ref{sec:storm}. Several simulations comparing the effects of different forcing conditions are also discussed. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

@@

We have used the Nucleus for European Modelling of the Ocean (NEMO) framework in its regional configuration \citep{madec2012nemo} to develop an ocean model for the Salish Sea. NEMO is a highly modular tool used for studying ocean physics, ocean-ice interactions, and the bio-geochemical properties of the ocean. NEMO's ocean core solves the three-dimensional hydrostatic equations of motion for an incompressible fluid under the Boussinesq approximation on a structured computational grid.  Although these features are not being used in the present work, NEMO's options for grid nesting and bio-geochemical coupling make it a useful tool for studying the complex physics and bio-geochemical interactions within the Strait of Georgia. This work focuses on the physical configuration of the Salish Sea model, in particular, determining appropriate forcing and boundary conditions for accurate reproduction of tidal amplitudes and phases as well as storm surge elevations. Future work will include bio-geochemical coupling and data assimilation. NEMO version 3.4 was used in this study.  NEMO has been successfully used in other coastal modelling studies in the Northeast Atlantic \citep{maraldi2013nemo} and the Mediterranean Sea \citep{brossier2012sensitivity}.}

The modelled domain extends from the Strait of Juan de Fuca to Puget Sound to Johnstone Strait (Figure \ref{fig:domain}). Bathymetry from the Cascadia physiography data set \citep{haugerud1999digital} was smoothed to limit large changes in depth across grid cells, so that $\Delta h/\bar{h} \leq 0.8$, where $\Delta h$ is the difference in depth between two adjacent grid cells and $\bar{h}$ is their average depth. For model stability, additional smoothing at the Strait of Juan de Fuca western boundary was imposed to achieve constant depth across the first ten grid cells inward from the open boundary. Depths between 0 m and 4 m were set to 4 m.  As depicted in Figure \ref{fig:domain}, the numerical grid is rotated $29^{\circ}$ counter-clockwise of North in order to maintain computational efficiency since currents within the Strait of Georgia are mainly aligned with this rotated axis.  A magnified image of the complex passages through the San Juan and Gulf Islands is provided in Figure \ref{fig:zoom}.

The curvilinear orthogonal numerical grid has been divided into 398 by 898 by 40 grid cells, which results in an almost uniform horizontal resolution with grid spacing approximately 440 m by 500 m. The 40 vertical $z$-levels were stretched gradually in order to achieve higher resolution in the surface layer, with 1 m vertical grid spacing down to about 10 m in depth. Below 10 m the grid was stretched to a maximum grid spacing of 27 m in the lowest layer. At the bottom boundary, partial $z$-levels were used in order to limit large changes in bathymetry across grid cells \citep{madec2012nemo}.  Model constraints require that all passages be at least two grid points wide.  This restriction required choices in either widening or closing narrow straits.  In particular, Active Pass was closed and Porlier Pass was widened (Figure \ref{fig:zoom}). All of these compromises are expected to have significant effects locally but minimal effects on the large scale.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{Figures/bathy.pdf}
\caption{ Model domain (highlighted area) including bathymetry, rivers (green circles), and storm surge locations of interest  (red stars). The black rectangle represents the region displayed in Figure \ref{fig:zoom}.}\label{fig:domain}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[scale=0.4]{Figures/bathy_zoom.pdf}
\caption{Model bathymetry and coastlines in the passages between the San Juan and Gulf Islands.}\label{fig:zoom}
\end{figure}

@@

Temperature and salinity at the Juan de Fuca boundary were taken from a weekly climatology which was created from results of a model covering the Salish Sea and the west coast of Vancouver Island \citep{massonfine2012}.  Their results, originally on $s$-levels were interpolated onto $z$-levels and then onto the NEMO horizontal grid to give a two-dimensional forcing that varies with time.  To prepare the climatology all years (1995-2008) were averaged and results, approximately every 15 days, were interpolated to a weekly climatology. The model is relaxed to the forced temperature and salinity over the 10 grid points (about 5 km) closest to the open boundaries, using a flow relaxation scheme \citep{engedahl1995use}.  In the north, the seasonal climatologies of \citet{thomson1981oceanography} were used in the same manner.

Sea surface height anomaly (or remote forcing) was set uniformly across the mouth of Juan de Fuca using values from the Tofino tide gauge \citep{DFOObservations} which is outside of, but close to, the western boundary of the model domain (Figure \ref{fig:domain}).  A monthly climatology was produced using daily averages from 2000-2010, binning them by month, averaging and setting the yearly mean to zero.  For the storm surge simulations, hourly variations in sea surface height were used.  These values are the Tofino tide gauge values with the tides removed. Tidal predictions were generated with the tidal analysis package called t\_tide \citep{pawlowicz2002classical} using a time series from the year of the storm surge. A harmonic analysis was performed for 67 constituents. In order to prevent over-predicted constituents due to high surge years, the time series was first filtered by applying a tide-killing Doodson filter \citep{parker2007NOAA} and periods with large non-tidal amplitudes were removed. Long period constituents such as Sa, Ssa, etc were not used in the tidal predictions since these constituents are often contaminated by non-tidal energy from seasonal meteorological trends. Further, constituents with a low signal to noise ratio were neglected.  The storm surge simulations also used the sea surface height anomaly from Port Hardy, forced at the northern boundary in Johnstone Strait.

 Although Tofino is not the closest tide gauge to the western boundary, we found that the surge conditions on the western coast of Vancouver Island and near the mouth of the Strait of Juan de Fuca are fairly uniform as summarized in Table \ref{tab:pacificsurge}. Generally, surge amplitudes at Tofino are slightly larger and earlier than Neah Bay and Bamfield. The remote forcing is applied uniformly across the open boundaries. Given the similarities between the surge amplitude and phase at Neah Bay and Bamfield, this seems to be a reasonable assumption. In support of this assumption, a geostrophic slope calculation with an estuarine velocity of 20 cm s$^{-1}$ would imply a difference of less than 5 cm between the northern and southern sides of the Strait of Juan de Fuca. In the future, this uniform specification of sea level at the western boundary will be relaxed by taking values from the solution of a large-scale forecasting model.}

\begin{table}[h]
\centering 
\tbl{}Average peak surge amplitudes and delay in peak timing from Tofino at several tide gauges on the Pacific coast. Averages obtained over the five surge events hindcasted in this article. A negative delay indicates that the peak surge occurred later than at Tofino.}
{\begin{tabular}{|l |c | c|}
\hline 
Location & Mean surge amplitude (m) & Mean delay from Tofino (hrs) \\
\hline
Port Hardy & 0.491 & -2.67\\
Winter Harbour & 0.572 & 0.333 \\
Tofino  & 0.707 & -- \\
Bamfield & 0.658 & -0.5 \\
Neah Bay & 0.658 & -1.83 \\
Toke Point & 1.17 & -2.83 \\
\hline
\end{tabular}}
\label{tab:pacificsurge}
\end{table}

The tidal forcing and  non-tidal remote forcing were applied using the NEMO Flather scheme \citep{flather1994storm, madec2012nemo}, a radiation scheme that allows disturbances to propagate out of the domain without reflections off of the open boundary. The barotropic velocity component of the remote forcing was set to zero. At the west, the baroclinic velocities at the boundary were set equal to the values inside the boundary (zero-gradient boundary conditions).  This scheme is not part of core NEMO 3.4 (the version used here).  Zero gradient conditions were chosen because the baroclinic velocity at the mouth of Juan de Fuca is primarily estuarine and thus set by density variations between inside and outside the domain.  At the northern boundary, we choose the `no boundary condition' of NEMO that sets up an estuarine circulation by relaxing the temperature and salinity over a width of 10 grids to the seasonal climatology.

@@

 The ocean surface is forced with momentum fluxes, solar and non-solar heat fluxes, and precipitation from a 33-km global atmospheric reforecasting model called CGRF suitable for use in ocean modelling \citep{smith2014new}. Forecasts from the period of 2002-2012 are available. One additional simulation employs the western Canada component of the High Resolution Deterministic Prediction System (HRDPS), a nested 2.5 km resolution atmospheric model provided by Environment Canada \citep{ECModel}. In addition to momentum and heat fluxes, forcing due to the surface atmospheric pressure is also included in the ocean momentum equations \citep{madec2012nemo}. In the open ocean, the surface atmospheric pressure primarily induces an inverse barometer (IB) sea surface height,
\begin{equation}
 \eta_{IB} = -\frac{1}{g\rho_{0}}\left(P_{s}-P_0\right), \label{eq:inverse}
\end{equation}
where $g$ is the acceleration due to gravity, $\rho_{0}$ is a reference density set to 1035 kg m$^{-3}$, $P_{s}$ is the sea level atmospheric pressure from the atmospheric model, and $P_0$ is a reference sea level pressure (set to 101,000 Pa). Since the atmospheric model employs a terrain following vertical coordinate the model surface pressure has been corrected to sea level using \citep{holton1992introduction}
\[ P_s = P_1\left(\gamma\frac{z_1}{T_1} +1 \right)^\frac{g}{\gamma R},\]
where $P_1$ and $T_1$ are the model pressure and temperature at height $z_1$, $R$ is the ideal gas constant, and $\gamma$ is a temperature lapse rate set to 0.0098 K m$^{-1}$. In coastal and shelf seas, the atmospheric pressure forcing can also induce significant non-IB sea surface height, as demonstrated in previous studies of the Hudson Bay and Labrador shelf system \citep{wright1987influence,young1995synoptic}.

River input provides a significant volume of fresh water to the Salish Sea and can influence stratification, circulation, and primary productivity. However, most rivers in the domain are not gauged so parametrizations were required to represent river flow. \citet{morrison2011rivers} provides a method for estimating freshwater runoff in the Salish Sea region based on precipitation. Monthly runoff volumes for each watershed for each year from 1970 to 2012 were acquired from \citet{morrison2011rivers}, as well as monthly averages. Freshwater runoff from each watershed was divided among the rivers in that watershed. The area drained by each river was estimated from Toporama maps by the Atlas of Canada and watershed maps available on the Washington State government website. The watersheds included in our model were Fraser (which represents approximately 44\% of the freshwater input into our domain), Skagit (12\%), East Vancouver Island (North and South) (12\%), Howe (7\%), Bute (7\%), Puget (6\%), Juan de Fuca (5\%), Jervis (4\%) and Toba (3\%). The monthly flow from each river was input as a point source from 0 to 3 metres at the model point closest to the mouth of each river. Incoming water was assumed to be fresh and at surface ocean temperature. A total of 150 rivers were parametrized by this method and their locations are indicated by the green dots in Figure \ref{fig:domain}.

Initial conditions for temperature and salinity came from a CTD cast in the middle Strait of Georgia taken in Sept 2002 \citep{pawlowiczetal2007}. Conditions were initially uniform horizontally and velocity was initialized at zero. The model was spun up for 15.5 months from the initial conditions above, starting Sept 16, 2002, using atmospheric forcing from 2002-2003, climatological temperature and salinity and sea surface height at the boundaries, with tides and climatological river output.  All storm surge runs were started at least three days prior to the event of interest with zero initial velocities and sea surface height and a stratification profile from model spin up.  The modelled sea surface height adjusted to forcing in less than one day as determined by comparing a simulation with restart conditions to one with velocities and sea surface height initialized to zero.

@@

In general, the K$_1$ tidal harmonics are well-represented by the model with complex differences below 5 cm at all stations except Seymour Narrows (station 27). Seymour Narrows is a very constricted region in Johnstone Strait and is not well resolved by the model. As such, the M$_2$ amplitude and phase are also poorly reproduced here.  The model's M$_2$ amphidrome, indicated by the drop in amplitude around station 5 (Victoria), is too strong resulting in lower than observed M$_2$ amplitudes in this region.  The model also represents the dip in M$_2$ amplitude around stations 25 and 26 reasonably well. Additionally, the M$_2$ phase errors in the Strait of Juan de Fuca (stations 1-7) are quite large, leading to large complex differences in this region.  The match in the Strait of Georgia (stations 10-24) is very good, with complex differences less than 6 cm, in part due to tidal tuning for highly accurate tides in the main region of interest. Because the tides were tuned for a good match in the Strait of Georgia, the match in the Strait of Juan de Fuca was compromised; the effect is particularly noticeable at the station closest to the open boundary (Port Renfrew, 1).  Details on the tidal tuning procedure are provided in Appendix \ref{sec:appendix}.

@@

\section{Storm Surge Hindcasts}\label{sec:storm}
A series of five storm surge hindcasts is presented next. Storm surge simulations were started from rest at least three days prior to the event of interest to allow for adjustment of model currents. Stratification conditions from model spin up at a time close to the event (within five days) were chosen to initialize the temperature and salinity fields. Boundary conditions include atmospheric forcing (momentum and heat fluxes and precipitation) and fresh water input from rivers at the surface, sea surface height anomaly from observations at the open boundary, and tidal forcing with eight tidal constituents.
